---
title: "Cornus transcriptome"
author: "Eugenia Nikonorova"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggpubr)
library(tidyverse)
library(dplyr)
library(paletteer)
library(stringr)
library(ggtree)
library(reshape2)
library(viridis)
library(readxl)
```


#1 Plotting BUSCO results 

```{r}
library(ggplot2)

busco <- read.csv(
  "Files/busco.csv",
  header = TRUE,
  dec = ".",
  sep = ";",
  stringsAsFactors = FALSE
)

busco$type <- factor(busco$type)
busco$db   <- factor(busco$db)

p <- ggplot(busco, aes(x = db, y = data, group = type)) +
  geom_bar(
    aes(fill = type),
    stat = "identity",
    position = position_dodge(width = 0.92),
    width = 0.85,
    color = "black",
    linewidth = 0.3
  ) +
  geom_text(
    aes(label = data),
    position = position_dodge(width = 0.92),
    vjust = -0.25,   # ← ключевое исправление
    size = 3.8,
    color = "black"
  ) +
  scale_fill_grey(start = 0.2, end = 0.9) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1))  # чтобы было место справа для чисел
  ) +
  labs(x = "", y = "", fill = "") +
  theme_classic() +
  theme(
    legend.position = c(0.98, 0.98),
    legend.justification = c(1, 1),
    legend.background = element_rect(fill = "white", color = NA),
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 14),
    axis.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

p

ggsave("plots/busco_statistics.png", plot = p, width = 10, height = 6, dpi = 300)
```


#2 Analyse GO annotation and create a figure

```{r, error=FALSE, warning=FALSE }

go_df <- read.csv("Files/GO_terms_highest_level_per_transcript.csv")

# Count number of genes per GO term and ontology 
go_summary <- go_df %>%
  group_by(Ontology, GO_name) %>%
  summarise(NumGenes = n_distinct(Transcript_ID), .groups = "drop")

# Optional: Keep only top N terms per ontology 
top_go <- go_summary %>%
  group_by(Ontology) %>%
  slice_max(order_by = NumGenes, n = 15) %>%  # <- can adjust N if needed
  ungroup()

# Simple plot GO terms by function  
ggplot(top_go, aes(x = reorder(GO_name, NumGenes), y = NumGenes, fill = Ontology)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  coord_flip() +
  labs(
    title = "Gene Function Classification (GO)",
    x = "GO Term",
    y = "Number of Genes"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    legend.position = "top"
  )

# Best variant (for paper and diploma ITMO)

ggdotchart(top_go, x = "GO_name", y = "NumGenes",
           color = "Ontology",                                # Color by groups
         #  palette = c("#00AFBB", "#E7B800", "#FC4E07"), # Custom color palette                      
           add = "segments",                             # Add segments from y = 0 to dots
           rotate = TRUE,                                # Rotate vertically
           group = "Ontology",
            label = top_go$NumGenes, 
           dot.size = 6,                                 # Large dot size
           font.label = list(color = "black", size = 9, 
                             vjust = 0.5),               # Adjust label parameters
           ggtheme = theme_pubr()                        # ggplot2 theme
           ) +
  labs(title = "Gene Function Classification (GO)",
          x = "",
          y = "") +
  theme(legend.position = c(0.7, 0.9))


ggsave("plots/GO_ontology_barplot.png", width = 10, height = 10, dpi = 300)


```
#3 Analyse KEGG pathway and create a figure

```{r, error=FALSE, warning=FALSE }

# Read KEGG analysis data
kegg_data <- readLines("Files/KEGG/KEGG_pathways.txt")

# Extract groups, descriptors, and counts
groups <- c()
descriptors <- c()
counts <- c()
current_group <- NA

for (line in kegg_data) {
  line <- str_replace_all(line, "[\u00A0\u200B\uFFFD]", " ")  # Replace special chars with regular spaces
  line <- str_squish(line)  # Remove extra spaces

  if (!str_detect(line, "\\(")) {
    current_group <- line
  } else {
    line_clean <- str_replace(line, "^\\d+\\s+", "")
    
    descriptor <- str_trim(str_extract(line_clean, "^.*(?=\\s\\()"))
    count <- as.numeric(str_extract(line_clean, "(?<=\\()\\d+(?=\\))"))
    
    groups <- c(groups, current_group)
    descriptors <- c(descriptors, descriptor)
    counts <- c(counts, count)
  }
}



# Create dataframe
kegg_df <- data.frame(Group = groups, Descriptor = descriptors, Count = counts)
kegg_df$Group <- as.factor(kegg_df$Group)
kegg_df$Descriptor <- as.factor(kegg_df$Descriptor)


# Top 30 Descriptors
top_kegg_df <- kegg_df %>% 
  arrange(desc(Count)) %>% 
  slice_head(n = 30)

ggplot(top_kegg_df, aes(x = reorder(Descriptor, Count), y = Count, fill = Group)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  theme_bw() +
  labs(title = "Top 30 KEGG Pathway Annotations",
       x = "KEGG Descriptor",
       y = "Number of Genes") +
  theme(axis.text.y = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        legend.title = element_blank(),
        legend.position = "right",
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

# Add the Class column based on row indices
kegg_df$Class <- NA

kegg_df$Class[1:146] <- "Metabolism"
kegg_df$Class[147:175] <- "Genetic Information Processing"
kegg_df$Class[176:212] <- "Environmental Information Processing"
kegg_df$Class[213:247] <- "Cellular Processes"
kegg_df$Class[248:336] <- "Organismal Systems"
kegg_df$Class[337:426] <- "Human Diseases"

# Verify
head(kegg_df, 150)
table(kegg_df$Class)

# Summarize counts by Class and Group
summary_df <- kegg_df %>%
  group_by(Class, Group) %>%
  summarise(Total_Count = sum(Count)) %>%
  ungroup()

# Order groups by Class and Total_Count
summary_df <- summary_df %>%
  group_by(Class) %>%
  arrange(Class, Total_Count) %>%
  mutate(Group = factor(Group, levels = unique(Group))) %>%
  ungroup()



ggdotchart(summary_df, x = "Group", y = "Total_Count",
           color = "Class",                                # Color by groups
          # palette = c("#00AFBB", "#E7B800", "#FC4E07"), # Custom color palett                      # Sort value in descending order
           add = "segments",                             # Add segments from y = 0 to dots
           rotate = TRUE,                                # Rotate vertically
           group = "Class",
            label = summary_df$Total_Count, 
           dot.size = 6,                                 # Large dot size
           font.label = list(color = "black", size = 9, 
                             vjust = 0.5),  
          legend.title = "", 
           ggtheme = theme_pubr()                        # ggplot2 theme
           ) +
  labs(title = "KEGG Pathway annotation",
          x = "",
          y = "") +
   theme(legend.position = c(0.7, 0.9))

ggsave("plots/KEGG_plot.png", width = 10, height = 10, dpi = 300)


#Total number of classes KEGG 

kegg_df %>%
  summarise(Total_Count = sum(Count)) %>%
  arrange(desc(Total_Count))
sum(kegg_df$Count, na.rm = TRUE)

```


#4 Sum of annotation results in general, from resulted best_hits, etc files

```{r}
# Create vectors
category <- c(
              "RefSeq",
              "UniProt",
              "NCBI NR",
              "KEGG", 
              "GO", 
              "Unannotated")

count <- c(26681, 19496, 26898, 12326, 12766, 2077)

# Create data frame
anno_Cser <- data.frame(Category = category, Count = count)


ggbarplot(anno_Cser , x = "Category", y = "Count", 
           ylab = "Number of transcripts",  xlab = "", title = "") +
  geom_text(aes(label=Count), color="black",
            position=position_dodge(width=0.9), vjust=-0.25, hjust=0.5, size=4.5) +
  theme(axis.text.y = element_text(size = 16),
        axis.text.x = element_text(size = 16),
        axis.title = element_text(size = 16),
        strip.text = element_text(size = 16)) +
  theme( legend.position = "none")

ggsave("plots/Annotation_statistics.png", width = 10, height = 6, dpi = 300)
```


#5 Count annotions in final gff3 file


```{r, error=FALSE, warning=FALSE}
#Install  package
library(reticulate)
py_install("pandas")
```

```{python}
#Count annotions in final gff3 file
import pandas as pd

annotated_gff = "Files/annotated_final.gff3"

counters = {
    'mRNA_total': 0,
    'RefSeq_best_hit': 0,
    'UniProt_best_hit': 0,
    'NR_best_hit': 0,
    'KEGG_KO': 0,
    'Ontology_term': 0,
    'any_protein_hit': 0
}

with open(annotated_gff, 'r') as infile:
    for line in infile:
        if line.startswith('#') or line.strip() == '':
            continue

        parts = line.strip().split('\t')
        feature_type = parts[2]
        attributes = parts[8]

        if feature_type != 'mRNA':
            continue

        counters['mRNA_total'] += 1

        attr_dict = {}
        for field in attributes.split(';'):
            if '=' in field:
                key, value = field.split('=', 1)
                attr_dict[key] = value.strip()

        for key in [
            'UniProt_best_hit',
            'RefSeq_best_hit',
            'NR_best_hit',
            'KEGG_KO',
            'Ontology_term'
        ]:
            if key in attr_dict and attr_dict[key] != '':
                counters[key] += 1

        if any(attr_dict.get(k) for k in ['UniProt_best_hit', 'RefSeq_best_hit', 'NR_best_hit']):
            counters['any_protein_hit'] += 1

print("mRNA annotation summary:")
for k, v in counters.items():
    print(f"{k}: {v}")

# save for plotting
gff3_annotation_res = pd.DataFrame({
      "Category": list(counters.keys()),
      "Count": list(counters.values())
  })


```

#6 Plot the counts of annotions in final gff3 file
```{r, error=FALSE, warning=FALSE}
library(reticulate)

# get Python object
#tmp_res <- py$gff3_annotation_res$to_dict("list")
tmp_res <- py$gff3_annotation_res
gff3_annotation_res <- as.data.frame(py_to_r(tmp_res))
new_line <- c("Unannotated", 2077)
gff3_annotation_res <- rbind(gff3_annotation_res, new_line)

gff3_annotation_res$Category <-  sub("_best_hit", "", gff3_annotation_res$Category)
gff3_annotation_res$Category <-  sub("KEGG_KO", "KEGG", gff3_annotation_res$Category)
gff3_annotation_res$Category <-  sub("NR", "NCBI NR", gff3_annotation_res$Category)
gff3_annotation_res$Category <-  sub("Ontology_term", "GO", gff3_annotation_res$Category)



# fix column types (important!)
gff3_annotation_res$Count <- as.numeric(gff3_annotation_res$Count)
gff3_annotation_res$Category <- as.character(gff3_annotation_res$Category)

ggbarplot(gff3_annotation_res[ c(2:6, 8),], x = "Category", y = "Count", 
           ylab = "Number of transcripts",  xlab = "", title = "") +
  geom_text(aes(label=Count), color="black",
            position=position_dodge(width=0.9), vjust=-0.25, hjust=0.5, size=4.5) +
  theme(axis.text.y = element_text(size = 16),
        axis.text.x = element_text(size = 16),
        axis.title = element_text(size = 16),
        strip.text = element_text(size = 16)) +
  theme( legend.position = "none")

ggsave("plots/Annotation_statistics_gff3.png", width = 10, height = 6, dpi = 300)
```


**DQD/SDH analysis**

#6 Sequense identity plot (from matrix after MAFFT alignment)

```{r}

labels <- c("CserDQD/SDH1.1", "CserDQD/SDH1.2", "CserDQD/SDH2",
            "CserDQD/SDH3", "CserDQD/SDH4", "CserDQD/SDH5")

# Data matrix of identity for nt sequenses (including gaps, Clustal Omega alignment)
data_nt <- matrix(c(
  100, 85, 33, 33, 35, 52,
  85, 100, 37, 33, 37, 43,
  33, 37, 100, 69, 71, 59,
  33, 33, 69, 100, 72, 60,
  35, 37, 71, 72, 100, 62,
  52, 43, 59, 60, 62, 100
), nrow = 6, byrow = TRUE)

rownames(data_nt) <- labels
colnames(data_nt) <- labels

# Data matrix of identity for prot sequenses (including gaps, Clustal Omega alignment)
data_pr <- matrix(c(
  100, 97, 44, 44, 38, 57,
  97, 100, 45, 45, 39, 58,
  44, 45, 100, 59, 70, 47,
  44, 45, 59, 100, 63, 47,
  38, 39, 70, 63, 100, 43,
  57, 58, 47, 47, 43, 100
), nrow = 6, byrow = TRUE)

rownames(data_pr) <- labels
colnames(data_pr) <- labels


# Melt data
#data_melt <- melt(data_nt) #or 
data_melt <- melt(data_pr)


# Save high-resolution figure
ggsave("sequence_identity_heatmap.png", p, width = 6, height = 5, dpi = 600)

# more light variant
library(scales)
# Plot
p <- ggplot(data_melt, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "grey85", linewidth = 0.4) +
  geom_text(aes(label = value), size = 3.2, fontface = "bold") +

  # LIGHT custom gradient
  scale_fill_gradientn(
  colours = c("#f7fbff", "#c6dbef", "#6b92d6", "#fddbc7", "#f4a582"),
  values = scales::rescale(c(10, 40, 60, 80, 100)),
  limits = c(0, 100),
  name = "Identity (%)"
) +

  coord_fixed() +

  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_blank(),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 9),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6)
  )

print(p)

# Save high-resolution figure
ggsave("plots/sequence_identity_heatmap.png", p, width = 6, height = 5, dpi = 600)

```

#7 Plots for Figure 2C, D - NWK tree as a dendrogram beside the heatmap 

```{r}

library(ape)
library(gplots)

# 1. Sequence identity matrix
# for nt - for AA change matrix and mwk tree
mat <- matrix(c(
 100, 97, 44, 44, 38, 57,
  97, 100, 45, 45, 39, 58,
  44, 45, 100, 59, 70, 47,
  44, 45, 59, 100, 63, 47,
  38, 39, 70, 63, 100, 43,
  57, 58, 47, 47, 43, 100
), nrow = 6, byrow = TRUE)

labels <- c(
  "CserDQD/SDH1.1",
  "CserDQD/SDH1.2",
  "CserDQD/SDH2",
  "CserDQD/SDH3",
  "CserDQD/SDH4",
  "CserDQD/SDH5"
)

rownames(mat) <- labels
colnames(mat) <- labels


nwk <- "(CserDQD/SDH1.1:0.00772569,CserDQD/SDH1.2:0.0211617,((CserDQD/SDH2:0.166335,(CserDQD/SDH3:0.248687,CserDQD/SDH4:0.208442):0.0965437):0.519644,CserDQD/SDH5:0.346317):0.429086);"

tree <- read.tree(text = nwk)

#plot(tree)

# Root the tree using one tip as outgroup
tree_rooted <- root(
  tree,
  outgroup = c("CserDQD/SDH1.1", "CserDQD/SDH1.2"),
  resolve.root = TRUE
)

# Optional check
print(is.rooted(tree_rooted))  # should be TRUE
#plot(tree_rooted) #Same as unrooted tree

# 3. Make branch lengths suitable for dendrogram conversion

tree_grafen <- compute.brlen(tree_rooted, method = "Grafen")

# Convert phylo -> hclust -> dendrogram
hc <- as.hclust(tree_grafen)
dend <- as.dendrogram(hc)

# Reorder matrix to match tree
tip_order <- hc$labels[hc$order]
mat_ord <- mat[tip_order, tip_order]


# 3. Mid-contrast publication palette

my_cols <- colorRampPalette(c(
  "#f7fbff", "#c6dbef", "#6b92d6", "#fddbc7", "#f4a582"
))(100)


# 4. Save as PNG

png("pr_sequence_identity_tree_heatmap.png", width = 2600, height = 1800, res = 300)


heatmap.2(
  mat_ord,
  Rowv = NA,
  Colv = dend,
  dendrogram = "column",
  trace = "none",
  key = F,
  key.title = "Identity",
  key.xlab = "%",
  density.info = "none",
  col = my_cols,
  breaks = seq(0, 100, length.out = 101),
  
  cellnote = mat_ord,   # add % sign
  notecol = "black",
  notecex = 1.0,
  font = 2,                          # <-- THIS makes them bold
  
  margins = c(12, 16),
  sepcolor = "grey85",
  colsep = 1:ncol(mat_ord),
  rowsep = 1:nrow(mat_ord),
  cexRow = 1.1,
  cexCol = 1.0,
  srtCol = 45,
  labRow = rownames(mat_ord),
  labCol = colnames(mat_ord)
)

dev.off() 
```


#9  Figure 3 - Sequence logo plots of cpTP- and non-cpTP amino acids (the first 80)

```{r}

if (!requireNamespace(c("tidyverse",
  "readxl",
  "Biostrings",
  "ggseqlogo",
  "patchwork"), quietly = TRUE)) install.packages(c("tidyverse",
  "readxl",
  "Biostrings",
  "ggseqlogo",
  "patchwork"))


library(tidyverse)
library(readxl)
library(Biostrings)
library(ggseqlogo)
library(patchwork)

# METADATA FOR cpTP presence (containd data on cpTP presence, QDH and GA-forming activity)

meta_file <- "Files/Alignment/metadata_fig_4.xlsx"
meta_raw <- read_xlsx(meta_file)

#rename columns for simplicity
colnames(meta_raw) <- c("seq_id", "gallate", "qdh", "cp")

meta <- meta_raw %>%
  mutate(
    gallate = as.numeric(gallate) > 0,
    qdh = as.numeric(qdh) > 0,
    cp = as.numeric(cp) > 0
  )

# input fasta with DQD/SDHs 

fasta_file <- "Files/Alignment/For_alignment.fasta"
faa <- readAAStringSet(fasta_file)

seq_df <- tibble(
  seq_id = names(faa),
  sequence = as.character(faa)
)

# clean IDs if FASTA headers contain descriptions
seq_df <- seq_df %>%
  mutate(seq_id = str_extract(seq_id, "^[^ ]+"))

# EXTRACT FIRST 80 AA

seq_df <- seq_df %>%
  mutate(first50 = str_sub(sequence, 1, 50))


# merge metadata and fasta

dat <- seq_df %>%
  inner_join(meta, by = "seq_id")


# SPLIT GROUPS

cp_yes <- dat %>%
  filter(cp == "TRUE") %>%
  pull(first50)

cp_no <- dat %>%
  filter(cp != "TRUE" | is.na(cp)) %>%
  pull(first50)




# PROBABILITY LOGOS


pp1 <- ggseqlogo(cp_yes, method = "prob") +
  ggtitle("Chloroplast transit peptide predicted") +
  labs(
    x = NULL,
    y = "Relative occurrence"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(size = 8),
    legend.position = "none"
  )

pp2 <- ggseqlogo(cp_no, method = "prob") +
  ggtitle("No chloroplast transit peptide") +
  labs(
    x = "Residue position",
    y = "Relative occurrence"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(size = 8),
    legend.position = "none"
  )


# BITS LOGOS

pp3 <- ggseqlogo(cp_yes, method = "bits") +
  ggtitle("Chloroplast transit peptide predicted") +
  labs(
    x = NULL,
    y = "Bits"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(size = 8),
    legend.position = "none"
  )

pp4 <- ggseqlogo(cp_no, method = "bits") +
  ggtitle("No chloroplast transit peptide") +
  labs(
    x = "Residue position",
    y = "Bits"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(size = 8),
    legend.position = "none"
  )


# COMBINE INTO ONE FIGURE


final_logo_plot <-
  (pp1 | pp3) /
  (pp2 | pp4) +
  plot_annotation(
    tag_levels = "A"
  )

print(final_logo_plot)


# SAVE


ggsave(
  "plots/cpTP_combined_sequence_logos.png",
  final_logo_plot,
  width = 20,
  height = 8,
  dpi = 300
)



```

#8A calculation of Hydrophobicity profile and Uncharged Amino Acids of cpTP
```{r}

library(tidyverse)
library(Biostrings)
library(ggseqlogo)
library(patchwork)
library(Peptides)
library(zoo)
library(readxl)


# Kyte-Doolittle hydrophobicity scale

hydro_scale <- c(
  A = 1.8,
  R = -4.5,
  N = -3.5,
  D = -3.5,
  C = 2.5,
  Q = -3.5,
  E = -3.5,
  G = -0.4,
  H = -3.2,
  I = 4.5,
  L = 3.8,
  K = -3.9,
  M = 1.9,
  F = 2.8,
  P = -1.6,
  S = -0.8,
  T = -0.7,
  W = -0.9,
  Y = -1.3,
  V = 4.2
)


# Hydrophobicity profile

calc_hydro <- function(seq) {

  aa <- strsplit(seq, "")[[1]]

  hydro <- hydro_scale[aa]

  tibble(
    pos = seq_along(hydro),
    hydro = as.numeric(hydro)
  )
}


# Residue frequency by position


count_residue <- function(seqs, residues) {

  split_seqs <- lapply(seqs, function(x)
    strsplit(x, "")[[1]])

  maxlen <- max(sapply(split_seqs, length))

  mat <- matrix(NA, nrow = length(split_seqs),
                ncol = maxlen)

  for(i in seq_along(split_seqs)) {
    mat[i, 1:length(split_seqs[[i]])] <- split_seqs[[i]]
  }

  freq <- apply(mat, 2, function(col) {
    mean(col %in% residues, na.rm = TRUE)
  })

  tibble(
    pos = seq_along(freq),
    freq = freq
  )
}

#Hydrophobicity analysis
hydro_yes <- map_df(cp_yes,
                    calc_hydro,
                    .id = "seq") %>%
  group_by(pos) %>%
  summarise(mean_hydro = mean(hydro, na.rm = TRUE)) %>%
  mutate(group = "cpTP")

hydro_no <- map_df(cp_no,
                   calc_hydro,
                   .id = "seq") %>%
  group_by(pos) %>%
  summarise(mean_hydro = mean(hydro, na.rm = TRUE)) %>%
  mutate(group = "non-cp")

hydro_all <- bind_rows(hydro_yes, hydro_no)

#Uncharged Amino Acids
unch_yes <- count_residue(cp_yes,
                         c("S", "T", "C", "N", "Q", "Y",
                           "G","A", "V", "L","I", "M", "F", "W", "P")) %>%
  mutate(group = "cpTP")

unch_no <- count_residue(cp_no,
                       c("S", "T", "C", "N", "Q", "Y",
                           "G","A", "V", "L","I", "M", "F", "W", "P")) %>%
  mutate(group = "non-cp")

unch_df <- bind_rows(unch_yes, unch_no)

```


#8B Plot the data of Hydrophobicity profile and Uncharged Amino Acids of cpTP

```{r}
pub_theme <- theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    #axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    plot.title = element_text(
      face = "bold",
      size = 12
    ),
    legend.title = element_blank(),
    legend.position = "top",
    legend.text = element_text(size = 12)
  )

# ==========================================
# COLORS
# ==========================================

my_cols <- c(
  "cpTP" = "#2E8B57",   # green
  "non-cp" = "#D95F02" # orange
)

# ==========================================
# PANEL A — HYDROPHOBICITY
# ==========================================

p_hydro <- ggplot(
  hydro_all,
  aes(
    x = pos,
    y = mean_hydro,
    color = group
  )
) +
  geom_smooth(
    se = FALSE,
    linewidth = 1.0,
    span = 0.3
  ) +
  scale_color_manual(values = my_cols) +
  scale_x_continuous(
    breaks = seq(0, 80, 10)
  ) +
  labs(
    x = "Amino acid position",
    y = "Mean hydrophobicity",
    title = "Hydrophobicity profile"
  ) +
  theme_classic() +
  pub_theme

# ==========================================
# PANEL B — UNCHARGED AA
# ==========================================

p_unch <- ggplot(
  unch_df,
  aes(
    x = pos,
    y = freq * 100,
    color = group
  )
) +
  geom_smooth(
    se = FALSE,
    linewidth = 1.0,
    span = 0.3
  ) +
  scale_color_manual(values = my_cols) +
  scale_x_continuous(
    breaks = seq(0, 80, 10)
  ) +
  labs(
    x = "Amino acid position",
    y = "Uncharged residues (%)",
    title = "Uncharged amino acids"
  ) +
  theme_classic() +
  pub_theme +
  theme(
    legend.position = "none"
  )

# ==========================================
# COMBINE
# ==========================================

final_plot <- p_hydro / p_unch +
  plot_annotation(
    title = "",
    theme = theme(
      plot.title = element_text(
        face = "bold",
        size = 20,
        hjust = 0.5
      )
    )
  )

final_plot

ggsave(
  "plots/cpTP_characteristics.png",
  final_plot,
  width = 10,
  height = 5,
  dpi = 600
)
```



#9 Figure 5 - use DQD/SDHs alignment, data from phylogenetic tree and metadata (cpTP presence, QDH and GA-forming activity)


```{r}

if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

# Install if needed
# devtools::install_github("YuLab-SMU/ggmsa")
# BiocManager::install(c("Biostrings"))
# install.packages(c("ggplot2", "readr"))

library(ggmsa)
library(Biostrings)
library(ggplot2)
library(readr)


# INPUT FILES

alignment_file <- "Files/Alignment/ClostalO.aln.fa"
tree_file <- "Files/Alignment/ClostalO.aln.fa.contree"

# Output files
reordered_fasta_out <- "Files/Alignment/ClostalO.aln.reordered.fa"
mapping_out <- "Files/Alignment/AtSDH_alignment_column_map.csv"
table_out <- "Files/Alignment/AtSDH_position_extraction_table.csv"

# Reference sequence name
reference_name <- "AtSDH"

# AtSDH residue positions of interest
domains <- list(
  DQD = c(124, 126, 128, 155, 214, 241, 269, 271, 279, 291, 304),
  SDH_substrate = c( 336, 338,  381, 385, 406, 407, 422, 423, 550, 578, 582),
  SDH_cofactor = c(483, 484, 485, 488, 522, 523)
)

# Number of visual spacer columns between blocks
spacer <- 3

# METADATA FOR DOTS (containd data on cpTP presence, QDH and GA-forming activity)

meta_file <- "Files/Alignment/metadata_fig_4.xlsx"
meta_raw <- read_xlsx(meta_file)

#rename columns for simplicity
colnames(meta_raw) <- c("seq_id", "gallate", "qdh", "cp")

meta <- meta_raw %>%
  mutate(
    gallate = as.numeric(gallate) > 0,
    qdh = as.numeric(qdh) > 0,
    cp = as.numeric(cp) > 0
  )

clade_map <- bind_rows(
  data.frame(seq_id = c("AtSDH", "BrapaQDH", "BnapusQDH"), clade = "I"),
  data.frame(seq_id = c("CserDQD/SDH1.1", "CserDQD/SDH1.2", "Poptr4", "EgQDH2"), clade = "II"),
  data.frame(seq_id = c("CserDQD/SDH5", "CsDQD/SDHb", "CasSDH1", "NtSDH2", "SlQDH",
                        "Poptr2", "Poptr3", "EcDQD/SDH4a", "EcDQD/SDH4b", "EgQDH1", "VvSDH2"), clade = "III"),
  data.frame(seq_id = c("CserDQD/SDH2", "CsDQD/SDHa", "NtSDH1", "SlSDH1", "VvSDH1",
                        "JrSDH", "FvSDH3", "FvSDH4", "Poptr1", "EcDQD/SDH1", "EgDQD/SDH1", "PgSDH1"), clade = "IV"),
  data.frame(seq_id = c("CserDQD/SDH3", "CsDQD/SDHc", "CasSDH2", "EcDQD/SDH3", "EgDQD/SDH3",
                        "PgSDH3_2", "FvSDH1", "PgSDH3a_1", "PgSDH3a_2", "PgSDH3_1", "VvSDH3"), clade = "V"),
  data.frame(seq_id = c("CserDQD/SDH4", "CsDQD/SDHd", "CasSDH3", "Poptr5", "FvSDH2",
                        "EcDQD/SDH2", "EgDQD/SDH2", "PgSDH4", "VvSDH4"), clade = "VI")
)

# HELPERS

read_tree_tip_order <- function(tree_file) {
  txt <- paste(readLines(tree_file, warn = FALSE), collapse = "")
  txt <- gsub("\\s+", "", txt)

  # remove branch lengths like :0.123
  txt2 <- gsub(":[0-9eE+\\.-]+", "", txt)

  # split Newick into tokens
  parts <- unlist(strsplit(txt2, "[\\(\\),;]"))
  tips <- parts[nzchar(parts)]

  # remove possible internal node labels that follow ')'
  # in most .contree files, after removing branch lengths, tips remain the true labels
  unique(tips)
}

build_reference_map <- function(aligned_seq, ref_name = "AtSDH") {
  chars <- strsplit(as.character(aligned_seq), "")[[1]]

  ref_pos <- integer()
  aln_col <- integer()

  aa_counter <- 0
  for (i in seq_along(chars)) {
    if (chars[i] != "-") {
      aa_counter <- aa_counter + 1
      ref_pos <- c(ref_pos, aa_counter)
      aln_col <- c(aln_col, i)
    }
  }

  data.frame(
    reference = ref_name,
    residue_number = ref_pos,
    alignment_column = aln_col,
    stringsAsFactors = FALSE
  )
}

extract_positions_by_reference <- function(seqs, ref_map, positions, group_name) {
  aln_cols <- ref_map$alignment_column[match(positions, ref_map$residue_number)]

  seq_chars <- as.character(seqs)

  residues_per_seq <- lapply(seq_chars, function(s) {
    substring(s, aln_cols, aln_cols)
  })

  residue_mat <- do.call(rbind, residues_per_seq)
  colnames(residue_mat) <- paste0("AA", positions)

  grouped <- apply(residue_mat, 1, paste0, collapse = "")

  out <- as.data.frame(residue_mat, stringsAsFactors = FALSE)
  out[[group_name]] <- grouped
  out
}

make_combined_subset <- function(seqs, ref_map, domains, spacer = 3, spacer_char = "-") {
  seq_chars <- as.character(seqs)
  seq_names <- names(seqs)

  combined_strings <- character(length(seq_chars))
  axis_labels <- character()
  spacer_positions <- integer()
  domain_ranges <- list()

  current_pos <- 0

  for (i in seq_along(domains)) {
    ref_positions <- domains[[i]]
    aln_cols <- ref_map$alignment_column[match(ref_positions, ref_map$residue_number)]

    block_strings <- vapply(seq_chars, function(s) {
      paste0(substring(s, aln_cols, aln_cols), collapse = "")
    }, character(1))

    if (i == 1) {
      combined_strings <- block_strings
      start_pos <- 1
    } else {
      spacer_string <- paste(rep(spacer_char, spacer), collapse = "")
      combined_strings <- paste0(combined_strings, spacer_string, block_strings)

      new_spacers <- (current_pos + 1):(current_pos + spacer)
      spacer_positions <- c(spacer_positions, new_spacers)

      axis_labels <- c(axis_labels, rep("", spacer))
      current_pos <- current_pos + spacer
      start_pos <- current_pos + 1
    }

    axis_labels <- c(axis_labels, as.character(ref_positions))
    end_pos <- start_pos + length(ref_positions) - 1
    domain_ranges[[names(domains)[i]]] <- c(start_pos, end_pos)

    current_pos <- end_pos
  }

  aa_set <- AAStringSet(combined_strings)
  names(aa_set) <- seq_names

  list(
    aa_set = aa_set,
    axis_labels = axis_labels,
    spacer_positions = spacer_positions,
    domain_ranges = domain_ranges
  )
}


# LOAD DATA

protein_sequences <- readAAStringSet(alignment_file)
tip_order <- read_tree_tip_order(tree_file)

# Keep only tree tips found in alignment, in tree order
tip_order_found <- tip_order[tip_order %in% names(protein_sequences)]

# Add any alignment sequences not found in the tree at the end
tip_order_missing <- setdiff(names(protein_sequences), tip_order_found)
final_order <- c(tip_order_found, tip_order_missing)

protein_sequences_reordered <- protein_sequences[match(final_order, names(protein_sequences))]

# Save reordered FASTA
writeXStringSet(protein_sequences_reordered, filepath = reordered_fasta_out)


# MAP AtSDH RESIDUE NUMBERING TO ALIGNMENT COLUMNS

if (!(reference_name %in% names(protein_sequences_reordered))) {
  stop(paste("Reference sequence", reference_name, "not found in alignment."))
}

ref_seq <- protein_sequences_reordered[[reference_name]]
ref_map <- build_reference_map(ref_seq, ref_name = reference_name)

write_csv(ref_map, mapping_out)

# EXTRACT RESIDUES FOR ALL SEQUENCES

all_positions <- unlist(domains, use.names = FALSE)
all_positions <- unique(all_positions)

extraction_df <- data.frame(
  Sequence = names(protein_sequences_reordered),
  stringsAsFactors = FALSE
)

for (group_name in names(domains)) {
  tmp <- extract_positions_by_reference(
    seqs = protein_sequences_reordered,
    ref_map = ref_map,
    positions = domains[[group_name]],
    group_name = group_name
  )

  aa_cols <- paste0("AA", domains[[group_name]])
  aa_cols <- aa_cols[!aa_cols %in% names(extraction_df)]

  if (length(aa_cols) > 0) {
    extraction_df[aa_cols] <- tmp[aa_cols]
  }
  extraction_df[[group_name]] <- tmp[[group_name]]
}

write_csv(extraction_df, table_out)


# MAKE PLOTTING SUBSET WITH VISUAL SPACERS

res <- make_combined_subset(
  seqs = protein_sequences_reordered,
  ref_map = ref_map,
  domains = domains,
  spacer = spacer,
  spacer_char = "-"
)

aa_set <- res$aa_set
writeXStringSet(aa_set, filepath = "Files/Alignment/combined_subset_alignment.fasta")


# PLOT basic ggmsa 

p <- ggmsa(
  aa_set,
  color = "Chemistry_AA",
  font = "DroidSansMono",
  char_width = 0.6,
  seq_name = TRUE
) +
  scale_x_continuous(
  breaks = seq_along(res$axis_labels),
  labels = NULL,
  position = "top"
) + 
  theme(
  axis.text.x = element_text(size = 14, angle = 90, vjust = 0.5, hjust = -1.6),
  axis.text.y = element_text(size = 14),
  axis.ticks.x = element_blank(),
  panel.grid = element_blank()
)


# paint spacer columns white so they look empty
for (pos in res$spacer_positions) {
  p <- p + annotate(
    "rect",
    xmin = pos - 0.5, xmax = pos + 0.5,
    ymin = -Inf, ymax = Inf,
    fill = "white",
    colour = NA
  )
}


# optional domain titles
domain_centers <- sapply(res$domain_ranges, mean)

p <- p +
  annotate("text", x = domain_centers["DQD"], y = length(aa_set) + 5,
           label = "DQD domain", size = 5) +
  annotate("text", x = domain_centers["SDH_substrate"], y = length(aa_set) + 5,
           label = "SDH domain 
(substrate)", size = 5) +
  annotate("text", x = domain_centers["SDH_cofactor"], y = length(aa_set) + 5,
           label = "SDH domain 
(cofactor)", size = 5) +
  coord_cartesian(clip = "off") +
  theme(
    plot.margin = margin(20, 20, 20, 20)
  )


highlight_numbers <- c(214, 241, 338, 385, 423, 483, 484, 485)
label_df <- data.frame(
  x = seq_along(res$axis_labels),
  label = res$axis_labels,
  stringsAsFactors = FALSE
)

label_df$is_highlight <- label_df$label %in% highlight_numbers

y_labels <- length(aa_set) + 2

p <- p +
  geom_text(
    data = label_df,
    aes(x = x, y = y_labels, label = label),
    inherit.aes = FALSE,
    angle = 90,
    size = 5,
    colour = ifelse(label_df$is_highlight, "red", "black"),
    fontface = ifelse(label_df$is_highlight, "bold", "plain")
  )


# get exact row y positions from ggmsa plot
gb <- ggplot_build(p)

row_y <- sort(unique(gb$data[[1]]$y), decreasing = TRUE)
if (length(row_y) != length(names(aa_set))) {
  row_y <- sort(unique(gb$data[[2]]$y), decreasing = TRUE)
}

row_map <- data.frame(
  seq_id = names(aa_set),
  y = row_y
)


#annotation
ann <- row_map %>%
  left_join(meta, by = "seq_id") %>%
  left_join(clade_map, by = "seq_id") %>%
  mutate(
    gallate = ifelse(is.na(gallate), FALSE, gallate),
    qdh = ifelse(is.na(qdh), FALSE, qdh),
    cp = ifelse(is.na(cp), FALSE, cp)
  )

clade_centers <- ann %>%
  filter(!is.na(clade)) %>%
  group_by(clade) %>%
  summarise(y = mean(y), .groups = "drop")

clade_bounds <- ann %>%
  filter(!is.na(clade)) %>%
  group_by(clade) %>%
  summarise(
    ymin = min(y) - 0.45,
    ymax = max(y) + 0.45,
    ymid = mean(y),
    .groups = "drop"
  )

# x positions for metadata columns, placed to the left of the sequence names
x_ga    <- -7.5
x_qdh   <- -5.5
x_cp    <- -3.5
x_clade <- -1.2
x_clade_bracket <- -1.8


p2 <- p +
  coord_cartesian(
  xlim = c(-7, length(res$axis_labels) + 0.5), #move 
    clip = "off"
  ) +
  theme(
    plot.margin = margin(20, 20, 20, 40)
  ) +
  annotate("text", x = x_ga,    y = max(ann$y) + 1.8, label = "GA",    fontface = "bold", size = 5) +
  annotate("text", x = x_qdh,   y = max(ann$y) + 1.8, label = "QDH",   fontface = "bold", size = 5) +
  annotate("text", x = x_cp,    y = max(ann$y) + 1.8, label = "Cp",    fontface = "bold", size = 5) +
  annotate("text", x = x_clade, y = max(ann$y) + 1.8, label = "Clade", fontface = "bold", size = 5) +
  geom_point(
    data = ann %>% filter(gallate),
    aes(x = x_ga, y = y),
    inherit.aes = FALSE,
    colour = "#E78AC3", size = 3.5
  ) +
  geom_point(
    data = ann %>% filter(qdh),
    aes(x = x_qdh, y = y),
    inherit.aes = FALSE,
    colour = "#4C78A8", size = 3.5
  ) +
  geom_point(
    data = ann %>% filter(cp),
    aes(x = x_cp, y = y),
    inherit.aes = FALSE,
    colour = "#4DAF4A", size = 3.5
  ) +
  geom_segment(
  data = clade_bounds,
  aes(x = x_clade_bracket, xend = x_clade_bracket, y = ymin, yend = ymax),
  inherit.aes = FALSE,
  linewidth = 0.5
) +
  geom_segment(
  data = clade_bounds,
  aes(x = x_clade_bracket, xend = x_clade_bracket + 0.22, y = ymax, yend = ymax),
  inherit.aes = FALSE,
  linewidth = 0.5
) +
  geom_segment(
  data = clade_bounds,
  aes(x = x_clade_bracket, xend = x_clade_bracket + 0.22, y = ymin, yend = ymin),
  inherit.aes = FALSE,
  linewidth = 0.5
) +
  geom_text(
  data = clade_bounds,
  aes(x = x_clade, y = ymid, label = clade),
  inherit.aes = FALSE,
  fontface = "bold",
  size = 4.2
)

print(p2)

ggsave("plots/Figure 4.png", p2,
       width = 12, height = 12, dpi = 300, bg = "white")


```


